Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-07 summary.tex
edits to caveats, including moving them above the summary.  Don't want to end on caveats and neglected effects, even if some of them are future looking.  minor tweaks elsewhere to match.

@@

\subsection{Caveats and outlooks} 
We highlight below some outstanding issues that should be addressed in
future works.

\subsubsection{Viscosity}\label{caveats_visc}
% Depending on where the VSI operates in a
% disk, additional effects may become important. For example, in the
% outer parts of a PPD, where we estimate the VSI 
% operates, disk self-gravity (SG) may play a role. Vertical
% SG may favor the VSI by increasing $|\p_z\Omega|$ through
% enhancing the density stratification 
% (Eq. \ref{vertical_shear}). %sg in pert? no because kx>>1. but maybe for kx~1

We have neglected viscosity. 
% while \citetalias{barker15} find that inclusion of viscosity is needed for 
% convergence when studying surface modes in radially global disk
% models. %but not body modes? 
% Viscosity may also affect the
% thermodynamic requirement for the VSI, because viscous heating
% needs to be self-consistently included in the energy equation. This
% should be investigated in future works.   
An important effect of viscosity is setting a minimum wavelength (or 
maximum wavenumber) for which the VSI is relevant. Disturbances with
large $\khat\equiv k_xH$ may be subject to viscous decay. Adopting the standard
precription for the kinematic viscosity $\nu = \alpha c_s H$
\citep{shakura73}, where $\alpha$ is the dimensionless viscosity
parameter, the viscous timescale for a perturbation lengthscale  
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha\khat^2\Omega $. 
Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/h\Omega$ implies 
$\khat^2\lesssim h/\alpha$. Inserting $h \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha\sim
10^{-5}$. \citepalias[These small viscosity values are required for
the VSI to operate, see][]{nelson13}.

%\citetalias{nelson13} find that a
%dimensionless viscosity of $\alpha\lesssim O(10^{-4})$ is required for the
%VSI.

The above argument justifies our focus on moderate
wavenumbers, $\khat=O(10)$, which precludes extremely small-scale
disturbances such as surface modes. Nevertheless, future calculations
should explicitly account for viscosity. In particular, viscous
heating may influence the thermodynamic dependence of the VSI.

% needs to be self-consistently included in the energy equation.

% It is of interest to quantify the effect of viscosity 
% on the linear VSI. This would help identify, more accurately, the
% appropriate regions in a protoplanetary disk that are sufficiently
% laminar to develop the VSI.  This problem is, however, complicated by
% the fact that proper inclusion of viscosity introduces a meridional
% flow in the equilibrium state, as well as viscous heating
% in the energy equation.

\subsubsection{Convective overstability}
We have focused on parameter regimes relevant to the VSI,
namely perturbations with small radial lengthscales 
($k_xH\gtrsim 1$) in disks with short thermal timescales
($t_c\Omega\ll 1$). This excludes the `convective overstability'
recently discussed by \cite{klahr14} and \cite{lyra14} in vertically
unstratified disk models.

The convective overstability corresponds to slowly growing epicyclic motions.
It is most effective for $t_c\Omega \sim 1$ and disturbances with
$|k_x|\ll |k_z|$, where $k_z$ is a vertical wavenumber. These
conditions are mutually exclusive with the VSI.  
The maximum growth rate for convective overstability is given by 
\begin{align}
  \sigma_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
\end{align}
\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
square of the radial buoyancy frequency
(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.  
%Evaluating Eq. \ref{max_conv_gen} at the midplane of our disk models,
%we find  
%\begin{align}
%  \sigma_\mathrm{conv,max} =
%  -\frac{ h^2\OmK}{4\gamma\Gamma}\left(p+q\right)
%  \left[\left(\gamma-1\right)p-q\right]. \label{max_conv}
%\end{align}
In fact, for the MMSN considered in this work, convective
overstability does not apply because 
$N_r^2>0$.

For more general disk structures convective overstability is
possible, in principle. The models adopted for our numerical study in
\S\ref{numerical} do satisfy $N_r^2<0$, but convective overstability
require perturbations with long radial lengthscales,
$k_xH\ll 1$, for which radially global disk models would be more 
self-consistent.

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects can be represented by a thermal
relaxation timescale $t_c$. For  
application to PPDs we connected $t_c$ to the disk
properties such as surface density, temperature and opacity, making
$t_c$ depend on position and the perturbation 
lengthscales. One might improve the realism further by generalizing
the opacity model adopted here to be density (and hence height)
dependent.  
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment of radiative diffusion. 
% done roughly in this work -
% a proper study 
% involves a Laplacian operator and additional boundary
% conditions must be supplemented. 
We will consider these improvements in a
follow-up study. %additional heat source

>>>>

\section{Caveats and Neglected Effects} 
%We highlight below some outstanding issues that should be addressed in
%future works.

\subsection{Viscosity}\label{caveats_visc}
% Depending on where the VSI operates in a
% disk, additional effects may become important. For example, in the
% outer parts of a PPD, where we estimate the VSI 
% operates, disk self-gravity (SG) may play a role. Vertical
% SG may favor the VSI by increasing $|\p_z\Omega|$ through
% enhancing the density stratification 
% (Eq. \ref{vertical_shear}). %sg in pert? no because kx>>1. but maybe for kx~1

Our neglect of viscosity is valid in a laminar disk, because molecular viscosity is so small.
However turbulence could act as an effective viscosity which preferentially damps 
small scale modes.   Since the goal of VSI is to drive turbulent transport,
the most relevant modes for sustaining the VSI should be able to overcome
turbulent damping.

To roughly estimate the wavelengths which are prone to damping, we
consider 
%An important effect of viscosity is setting a minimum wavelength (or 
%maximum wavenumber) for which the VSI is relevant. Disturbances with
%large $\khat\equiv k_xH$ may be subject to viscous decay. Adopting 
the standard
precription for the kinematic viscosity $\nu = \alpha c_s H$
\citep{shakura73}, in terms of the dimsionless $\alpha$ 
parameter. The viscous timescale for a perturbation lengthscale  
$l\sim 1/k_x = H/\khat$ is $t_\mathrm{visc} = 1/\alpha\khat^2\Omega $. 
For significant growth, $t_\mathrm{visc}$ should be longer than the
characterstic VSI growth timescale $1/(h|q|\Omega)$, i.e.\ 
$\khat^2\lesssim |q|h/\alpha$. With $h \sim 0.05$, $q = -1$  and
 $\alpha$ from simulations \citepalias{nelson13},
we estimate that growth requires $\khat\lesssim 20$ for
$\alpha\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha\sim
10^{-5}$.

This argument justifies our focus on moderate
wavenumbers, $\khat=O(10)$. Future work should consider
the viscous effects in more detail, to better understand
how the VSI operates in nature and in simulations.

%suggestion: Keep this idea to yourself, since you don't have a result yet.
% In particular, viscous
%heating may influence the thermodynamic dependence of the VSI.

\subsection{Convective overstability}
The VSI bears some similarity to the `convective overstability'
 (CONO, \citealp{klahr14, lyra14}).   Both instabilities rely on thermal relaxation to 
overcome the stability of adiabatic perturbations, i.e.\ the Solberg-Hoiland
criteria (see \S\ref{solberg}).

The key differences are that the CONO neglects vertical buoyancy
and requires an imaginary horizontal buoyancy,
\begin{align}\label{Nr}
N_r^2 \equiv -{1 \over \rho C_P}{\p P \over \p r} {\p S \over \p r} < 0\, .
\end{align}
Since vertical buoyancy is an important stabilitizing influnce for the VSI, it should
be considered in future studies of CONO.

The sign of $N_r^2$ depends on disk parameters.  To demonstrate that the required parameters 
are somewhat extreme, we consider the midplane of vertically isothermal disks with 
$\Sigma \propto r^n$ so that $n = p+q/2 + 3/2$. Using Eqs. \ref{dSdr} and \ref{Nr} with $\gamma = 1.4$,
$N_r^2 < 0$ requires $n> 3(1/2 + q)$.  For standard disk temperature laws this requirement implies
a flat or rising surface density profile, e.g.\ $n = 3/14$ for $q = -3/7$ or $n = 0$ for $q = -1/2$.
Thus CONO is most like to operate at special locations, like the outside edges of 
disk gaps or holes and/or shadowed regions where $q < -1/2$.

Perhaps due to these physical differences, CONO operates in different regimes of parameter
space than the VSI.  CONO grows best for longer wavelengths $\khat \gtrsim 1$ and longer 
cooling times $\beta \sim 1$.

Future work should aim to understand these related instabilities in the same framework, thereby 
explaining the key differences.

%We have focused on parameter regimes relevant to the VSI,
%namely perturbations with small radial lengthscales 
%($k_xH\gtrsim 1$) in disks with short thermal timescales
%($t_c\Omega\ll 1$). This excludes the `convective overstability'
%recently discussed by \cite{klahr14} and \cite{lyra14} in vertically
%unstratified disk models.   
%
%The convective overstability corresponds to slowly growing epicyclic motions.
%It is most effective for $t_c\Omega \sim 1$ and disturbances with
%$|k_x|\ll |k_z|$, where $k_z$ is a vertical wavenumber. These
%conditions are mutually exclusive with the VSI.  
%The maximum growth rate for convective overstability is given by 
%\begin{align}
%  \sigma_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
%\end{align}
%\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
%square of the radial buoyancy frequency
%(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.  
%%Evaluating Eq. \ref{max_conv_gen} at the midplane of our disk models,
%%we find  
%%\begin{align}
%%  \sigma_\mathrm{conv,max} =
%%  -\frac{ h^2\OmK}{4\gamma\Gamma}\left(p+q\right)
%%  \left[\left(\gamma-1\right)p-q\right]. \label{max_conv}
%%\end{align}
%In fact, for the MMSN considered in this work, convective
%overstability does not apply because 
%$N_r^2>0$.   
%
%For more general disk structures convective overstability is
%possible, in principle. The models adopted for our numerical study in
%\S\ref{numerical} do satisfy $N_r^2<0$, but convective overstability
%require perturbations with long radial lengthscales,
%$k_xH\ll 1$, for which radially global disk models would be more 
%self-consistent.

\subsubsection{Radiative transfer} 
Our relatively simple treatment of cooling with an idealized dust opacity
could certainly be generalized in future works.  In hotter disks, gas phase opacities
must be considered.   In cold disks there are many choices for a dust opacity which 
varies with grain sizes and compositions. Dust properties could change the viability 
of the VSI for better or worse.  The radiative transfer itself
could be calculated with higher levels of sophistication, as is already being done in
numerical simulations \citep{stoll14}.

@@
